\documentclass[12pt]{article}

%% preamble: Keep it clean; only include those you need
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\usepackage[margin = 1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}
\usepackage{ragged2e}
\usetikzlibrary{positioning, arrows.meta}

\usepackage{listings}
\usepackage{xcolor}

\lstset{
  language=R,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue},
  commentstyle=\color{green!50!black},
  stringstyle=\color{orange},
  breaklines=true,
  showstringspaces=false
}

%% for double spacing
\usepackage{setspace}

% for space filling
\usepackage{lipsum}
% highlighting hyper links
\usepackage[colorlinks=true, citecolor=blue]{hyperref}

% Get line numbers for ease of referencing for the reviewers
\usepackage[]{lineno}
\linenumbers*[1]
% patches to make lineno work better with amsmath
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
        \expandafter\let\csname old#1\expandafter\endcsname\csname 
        #1\endcsname
        \expandafter\let\csname oldend#1\expandafter\endcsname\csname 
        end#1\endcsname
        \renewenvironment{#1}%
        {\linenomath\csname old#1\endcsname}%
        {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
        \patchAmsMathEnvironmentForLineno{#1}%
        \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
        \patchBothAmsMathEnvironmentsForLineno{equation}%
        \patchBothAmsMathEnvironmentsForLineno{align}%
        \patchBothAmsMathEnvironmentsForLineno{flalign}%
        \patchBothAmsMathEnvironmentsForLineno{alignat}%
        \patchBothAmsMathEnvironmentsForLineno{gather}%
        \patchBothAmsMathEnvironmentsForLineno{multline}%
}

\newtheorem{prop}{Proposition}
\newtheorem{thm}{Theorem}
\newtheorem{lem}{Lemma}

\theoremstyle{definition}
\newtheorem{dfn}{Definition}
\newtheorem{asp}{Assumption}
\newtheorem{ex}{Example}

%% meta data

\title{The Effect of Poverty on Cardiovascular Disease Mortality (Midterm Report)}
\author{Mathew Chandy\\
  Department of Statistics\\
  University of California, Los Angeles
}

\begin{document}
\maketitle


\section{Introduction}
\label{sec:intro}
My goal for this project is to investigate the causal
relationship between poverty rate and
cardiovascular disease (CVD) mortality rate at the county level. I am using 2016 data from several sources which I merge
by county FIPS code.
The first is the CDC Wonder dataset, which contains
information on age-adjusted CVD mortality rate ($Y$) for counties across the U.S. We use the age-adjusted outcome,
which standardizes to the year-2000 national population, to remove the mechanical effect of county age structure on mortality rates \citep{anderson1998age}.
The second is the American Community Survey, which contains
information on poverty rate ($X$), non-white population ($C$), which serves as a proxy for general racial composition, and \% aged 65 years older ($A$) 
which we retained as a covariate despite the age adjustment because standardization to the year-2000 reference population does not fully eliminate residual age confounding in counties with extreme demographic profiles.
We use the Small Area Health Insurance Estimates (SAHIE)
to obtain uninsurance rate ($M$) and the National Center for Health Statistics (NCHS) to obtain the urbanicity classification ($R$) as defined in 2013.  The urban-rural classification scheme has six ordinal levels: 
% https://www.cdc.gov/nchs/data/data-analysis/2023-File-Documentation-final.pdf. 
We obtain each county's medicaid expansion status ($G$) as of January 1st, 2016 from KFF.

\section{Causal Model}

Let us write out our causal model, where each county is an observational unit.
Let $\mathbf V = \{X, Y, A, C, M, G, R\}$, and
let $\mathbf  U = \{U_A, U_C,U_G,U_R\}$,
where $A \leftarrow U_A$, $C \leftarrow U_C$,
$G \leftarrow U_G$, and $R \leftarrow U_R$.
Here is a diagram of our assumed causal relationship
model:

\centering
\begin{tikzpicture}[
  node distance = 2cm and 2.5cm,
  every node/.style = {draw, circle, minimum size=0.9cm, font=\small},
  arr/.style = {-{Stealth}, thick}
]


\node (R) {$R$};
\node (X) [below left=of R]  {$X$};
\node (C) [below=of X]       {$C$};
\node (M) [below right=of X] {$M$};
\node (G) [right=of M]       {$G$};
\node (Y) [below right=of R] {$Y$};
\node (A) [above=of Y]       {$A$};

\draw[arr] (X) -- (Y);   % direct effect
\draw[arr] (X) -- (M);   % X -> M
\draw[arr] (M) -- (Y);   % mediated path
\draw[arr] (C) -- (X);   % C -> X
\draw[arr] (C) -- (M);   % C -> M (collider arrow)
\draw[arr] (C) -- (Y);   % C -> Y
\draw[arr] (R) -- (X);   % R -> X
\draw[arr] (R) -- (Y);   % R -> Y
\draw[arr] (R) -- (M);   % R -> M (added: rural uninsurance/access)
\draw[arr] (A) -- (Y);   % age -> Y only
\draw[arr] (G) -- (M);   % Medicaid -> M
\draw[arr] (G) -- (Y);   % Medicaid -> Y (direct, per DiD evidence)
\draw[<->, dashed, bend left=80] (C) to (R);
\draw[<->, dashed, bend left=80] (R) to (A);

\end{tikzpicture}

\justifying

Here is a short review of literature concerning
the causal relationships claimed in our causal diagram.
Poverty directly increases CVD mortality through stress,
malnutrition, and poor housing pathways independent of uninsurance \citep{kim2025mediating}. Poverty also reduces
healthcare access by pricing workers out of employer-sponsored and individual-market coverage, confirmed by the Affordable Care Act (ACA) Medicaid expansion difference-in-differences literature \citep{khatana2019association}, which also supports medicaid expansion as a cause of reduction in both uninsurance and CVD mortality directly and higher uninsurance rates as a cause of increase in CVD mortality. These conclusions are also supported by \citet{miller2021medicaid}, which links Medicaid expansion
to a 9.4\% reduction in annual mortality using individual-level linked survey and death records. Racial composition increases county poverty through structural racism operating via historically accumulated barriers in housing, employment, and lending \citep{bailey2021structural}. Racial composition also independently raises uninsurance rates - disparities persist within income an deducation groups, confirming this edge is not fully mediated by poverty \citep{lee2021convergence}. Lastly, racial composition increases CVD mortality independently of the poverty and uninsurance pathways, with \citet{geronimus2006weathering}
documenting widening racial gaps in biological stress accumulation that are not explained by income. Rural counties typically have higher poverty due to fewer labor markets and potential industrial decline \citep{farrigan2026rural} and higher CVD mortality through
hospital closures and scarcity of cardiologists that may be independent of poverty \citep{loccoh2022rural}.
The level of urbanicity also raises uninsurance rates through lower employer-sponsored coverage availability
and provider network shortages, though the raw uninsurance gap is smaller than the access gap, so this edge is partially but imperfectly captured by the SAHIE uninsurance measure. Finally, how old the population of a county is predicts CVD mortality as a precision variable. We also include a bidirected edge between urbanicity and non-white-population, with racial minorities making up about 22 percent of the American rural population in 2018, and 43 percent of the urban population \citep{castillo2020racial},
and one between urbanicity and county age, supported a trend of ageing in rural counties \citep{monnat2025us, cohen2023aging}.

\section{Queries of Interest}


Many variables in this study, including the treatment and the response, are continuous. This necessitates slightly
different notation than in the binary case. Instead of considering two levels $x_0$ and $x_1$ of the treatment poverty rate, we consider some level $x_0$ and 
$x_1 = x_0 + \delta$, where $\delta$ represents some change in the state of $X$.
For instance,
the total variation can be stated as
$$TV_{x_1,x_0}(y)=\mathbb E[Y|x_1]-E[Y|x_0]$$

I am interested in isolating the direct effect,
indirect effect, and spurious effect. If county poverty increases by $\delta$ percentage points:


\begin{enumerate}
    \item How much does age-adjusted CVD mortality change,
allowing all pathways (healthcare access + stress + structural factors) to operate?
    $$TE_{x_1,x_0}(y) = \mathbb E[Y_{x_1}]-\mathbb E[Y_{x_0}]$$
    \item How much does age-adjusted CVD mortality change,
isolating pathways (stress, housing, nutrition, structural racism channels, etc.) that do not operate through reduced healthcare?
$$NDE_{x_0,x_1}(y) = \mathbb E[Y_{x_1, M_{x_0}}]-\mathbb E[Y_{x_0}]$$
    \item How much does age-adjusted CVD mortality change,
isolating the indirect pathway through uninsurance?
$$NIE_{x_1,x_0}(y) = \mathbb E[Y_{x_1, M_{x_0}}]-\mathbb E[Y_{x_1}]$$
    \item How much of the poverty–mortality association is driven by
racial composition,
rurality,
Medicaid expansion, and 
age structure
rather than poverty itself?
    $$SE_{x_1,x_0}=\mathbb E[Y_{x_1}|x_0] - E[Y_{x_1}|x_1]$$
\end{enumerate}
To make notation a little simpler, we will refer to these effects without their subscripts, $TV_{x_1,x_0}(y)$ with $TV$, for example.
 


\section{Methods of Identification and Estimation}

To identify the total effect of $X$ on $Y$, we can use the backdoor adjustment set to identify $\mathbf Z= \{C,R\}$, which blocks the back-door paths $X\leftarrow C \rightarrow Y$, $X\leftarrow C \rightarrow M\rightarrow Y$, $X \leftarrow R \rightarrow Y$, and 
$X \leftarrow R \rightarrow M\rightarrow Y$. 
While $\{C,R\}$ constitutes the minimal adjustment set
for identifying $TE$, I utilize the expanded set $\mathbf W = \{C,R,A,G\}$ as this is an 
adjustment set sufficient to identify the $NDE$ and $NIE$. The addition of $G$ blocks
the back-door path $M\leftarrow G\rightarrow Y$. Although $A$ is not part of the minimal adjustment set necessary for $NDE$ identification, it is included as a precision variable
for reasons that will be discussed further.




% \begin{proof}

% Then we have,
% \[
% \begin{array}{r l l}
% f(Y_{x_1,M_{x_0}}=y) &= \int_m f(Y_{x_1, m}=y,M_{x_0}=m)dm & \text{(CUT)} \\
% &= \int_m f(Y_{x_1, m}=y,M_{x_0}=m)dm & \text{(Chain rule)} \\
% &= Q & \rightarrow\text{E }(1,2)
% \end{array}
% \]
% \end{proof}

Sketch? More flexible than PLM
\begin{enumerate}
    \item Train models with crossfitting (outcome model like XGBoost and density model like ?) on each fold
    \item Debias with influence function on out of fold data
    \item Average
    \item Influence function will have to be shown using strategy 2
    \item Decompose functionals at each data point 
    \item Compute $TV$ with nonparametric regression?
\end{enumerate}

Because of standard causal relationships $TE = NDE - NIE$, and $TV = TE - SE$, 
the minimal procedure requires us to only estimate $TV$, $TE$, and $NDE$. An estimate of the total variation $\widehat {TV}$ can be obtained
with naive nonparametric regression on $X$ alone without adjusting for any other covariates. In our application, we used local linear regression
with bandwidth optimized at 0.036.
% Assuming a linear relationship, $TV$ would be represented by $\beta_{TV}$ below:
% $$Y=\beta_0 + \beta_{TV} X + \epsilon,$$
% and could be estimated with $\hat \beta_{TV}$ from ordinary least squares. 
An estimate of the total effect $\widehat {TE}$ is slightly more complicated in the continuous setting. The target functional is $$\psi= \mathbb E \{ \mathbb E [Y|\mathbf W,X=x]\},$$ where we denote $\mu_x(\mathbf W) = \mathbb E[Y|\mathbf W, X=x]$ for $\mu_x(\mathbf w) = \mathbb E[Y|\mathbf W = \mathbf w,X=x]$. We make use of Strategy 2 from \citet{kennedy2017non}, which requires the following tricks:
\begin{enumerate}
    \item Pretend the data are discrete. In notation:
    $\psi\approx\sum_\mathbf{w} \mu_x(\mathbf w)p(\mathbf w)$, where $p(\mathbf w) = \mathbb P(\mathbf W = \mathbf w)$.
    \item We introduce an operator $\mathbb{IF}:\Psi\rightarrow L_2(\mathbb P)$ that maps functions $\psi: \mathcal P \rightarrow \mathbb R$ to their influence functions in a nonparametric setting where influence functions are treated as derivatives, enabling us to use 
    \begin{itemize}
        \item Product rule: $\mathbb {IF}(\psi_1\psi_2)=\mathbb {IF}(\psi_1)\psi_2+\psi_1\mathbb {IF}(\psi_2)$
        \item Chain rule: $\mathbb {IF}(f(\psi))=f'(\psi)\mathbb {IF}(\psi)$
    \end{itemize}
    \item Use influence function building blocks, for example:
    \begin{itemize}
        \item $\mathbb {IF}(\mu_x(\mathbf w))=\frac{\mathbf 1(\mathbf W=\mathbf w)}{\mathbb P(\mathbf W=\mathbf w)}\{Y-\mathbf E[Y|\mathbf W = \mathbf w]\}$.
    \end{itemize}
    
\end{enumerate}
Then we can obtain
\begin{align*}
    \mathbb{IF}(\psi)&=\mathbb{IF}\{\sum_\mathbf{w}\mu_x(\mathbf w)p(\mathbf w)\}=\sum_\mathbf{w}[\mathbb{IF}\{\mu_x(\mathbf w)\}p(\mathbf w) +\mu_x(\mathbf w)\mathbb {IF}\{p(\mathbf w)\}]\\
    &=Something SomethingSomething\\
    &=\frac{\mathbf 1(X=x_0)}{\pi(\mathbf W)}\{Y-\mu_x(\mathbf W)\}+\mu_x(\mathbf W)-\psi
\end{align*}

We will now demonstrate a similar estimate of the natural direct effect $\widehat {NDE}$, 
where the target functional is
$$$$

To estimate the causal parameters, I extend the Partially Linear Model (PLM) framework using Double Machine Learning (DML) as developed by \citet{chernozhukov2018double} to accommodate mediation. 
In addition to the outcome equation, I estimate a mediator equation for uninsurance. The structural system is
$$M = \alpha X + h(\mathbf W) + \nu$$
$$Y = \theta X + \gamma M + f(\mathbf W) + \epsilon$$
$$X = g(\mathbf W) + \eta,$$

where $\mathbf W$ denotes the set of observed pre-treatment confounders, $h(\cdot)$ and $f(\cdot)$
are unknown nuisance functions learned via machine learning, and $(\alpha, \theta, \gamma)$ are low-dimensional causal parameters. Here, $\alpha$ captures the causal effect of poverty on uninsurance, $\gamma$ captures the effect of uninsurance on CVD mortality holding poverty fixed, and $\theta$ captures the direct effect of poverty on mortality not operating through healthcare access.

This approach uses the \citet{robinson1988root} transformation to partial out the
effect of county-level confounders $\mathbf W$ by training ML algorithms to predict the CVD mortality rate, the poverty rate, and the uninsurance rate based on $\mathbf W$, and subsequently regressing the mortality residuals on the poverty residuals and uninsurance residuals.

The steps are
\begin{enumerate}
    \item Divide counties into $K$ folds
    \item For each fold $k$, train on the remaining $K - 1$ folds:
    \begin{itemize}
        \item Fit $\hat f^{(-k)}$: regress $Y$ on $\mathbf W$ (no $X$)
        \item Fit $\hat g^{(-k)}$: regress $X$ on $\mathbf W$ 
        \item Fit $\hat h^{(-k)}$: regress $M$ on $\mathbf W$ (no $X$)
    \end{itemize}
    
    \item Calculate residuals on held-out fold 
    $$\tilde Y=Y- \hat f^{(-k)}(\mathbf W)$$, 
    $$\tilde X=X- \hat g^{(-k)}(\mathbf W)$$,
    $$\tilde M=M- \hat h^{(-k)}(\mathbf W)$$
    \item Pool residuals across folds
    \item Regress $\tilde M$ on $\tilde X$ to identify $\hat \alpha$
    \item Regress $\tilde Y$ on $\tilde X$ and $\tilde M$ to identify 
    $\hat \theta$ and $\hat \gamma$
\end{enumerate}

Then $\widehat {TE}_{x_1,x_0}(y) = \hat \theta + \hat \gamma \hat \alpha$, $\widehat {NDE}_{x_0,x_1}(y) = \hat \theta$, $\widehat {NIE}_{x_1,x_0}(y) = -\hat \gamma \hat \alpha$. 


% Data
% https://wonder.cdc.gov/cmf-icd10.html
% tidycensus



    






\bibliography{main}
\bibliographystyle{chicago}

\end{document}